---
title: "Accessing elements"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This chapter looks at how to most easily access your data once they are in R objects.  

Note this uses "Base R" but you may prefer using `dplyr()` when workign with data.frames, which does similar in an (arguably) easier to understand syntax.  

But, it's worth knowing how to do this without relying on a package if possible.

## Subsetting using [ ]

Most R objects can have their individual elements reached via their numeric position.  These can be reached by using square brackets `[ ]`

```{r}
a_vector <- letters
a_vector[1]
a_vector[5]
a_vector[1:5]
a_vector[c(1,5)]
```

But when in a `data.frame` you also have to worry about the second dimension.

In this case, the `[ ]` notation is extended to include a comma: `[ , ]`.

* The position before the `,` indicates which row
* The position after the `,` indicates which column

_Note: This is sort of like R1C1 notation in Excel...except with a comma!_

```{r}
my_data <- mtcars

## the second row
my_data[2, ]

## the first column
my_data[ , 1]

## the second row, first colum
my_data[2, 1]

## the second and third row, and first and forth colum
my_data[2:3, c(1,4)]

```

## Character subsetting

Subsetting by numbers assumes the rows and columns are in the same order, which can be dangerous.  Safer is to use the name, if you know it:

```{r}
## First 5 rows of the "mpg" column
my_data[1:5, "mpg"]
```

You can aso use multiple columns:

```{r}
my_data[1:5, c("mpg","gear")]
```

And reorder or repeat columns (but it will rename them to avoid clashes via `make.names()`)

```{r}
my_data[1:5, c("gear","mpg", "mpg")]
```

Note that if you subset `lists()` or `data.frames()` with `[ ]` it will, by default, return a list or data.frame back.  

If you want to instead return the column vector, then use `[[ ]]` which returns whats in the list/data.frame column.

_This is confusing topic. It's right up there with `StringsAsFactors = FALSE`. This is where the console comes in handy when you're trying to make sure you have your syntax correct._

```{r}
my_data[["mpg"]]
```

### Lookup Tables

Character subsetting is how the lookup examples work, the only differences being that, this time, it's not a `data.frame` but a `vector` that is named using the `names()` function.

```{r}
## example with some user Ids
lookup <- c("Bill","Ben","Sue","Linda","Gerry")
names(lookup) <- c("1231","2323","5353","3434","9999")
lookup

## this is a big vector of Ids you want to lookup
big_list_of_ids <- c("2323","2323","3434","9999","9999","1231","5353","9999","2323","1231","9999")

## subset lookup with repeating columns by your data of ids
lookup[big_list_of_ids]
```

## $ operator

You can also find columns via the `$` operator on lists and data.frames:

```{r}
my_data$mpg
```

This `$` is a shortcut to subsetting via a character name:

```{r}
my_data[["mpg"]]
```

You can then apply subsetting to the result to get a specific number(s):

```{r}
## first 5 elements of mpg column
my_data$mpg[1:5]
```

## Subsetting using logical

You can also subset using `TRUE` and `FALSE`.  This is a good way to select rows.

For instance, to select all rows that are over 24 in the `mpg` column of `mtcars`.

We first construct the logical vector:

```{r}
## we first make a TRUE or FALSE vector for every mpg element over 24
over_24 <- mtcars$mpg > 24
over_24
```

We can then pass this into the row selector for mycars:

```{r}
## all rows over 24 and all columns
mtcars[over_24, ]
```

This is often shortend to one line:

```{r}
mtcars[mtcars$mpg > 24, ]
```

And if you only wanted certain columns, you can add that to the line too:

```{r}
mtcars[mtcars$mpg > 24, c("mpg","wt")]
```

These can start to look pretty confusing, but, once you get comfortable with the basic syntax, you will see how things break down. And, it can be useful to build up the final syntax iteratively, much as was done in the example above.

A couple of additional notes on the conditional selections (the use of `>` above):

* To set "equals to," use a double equals sign: `==`
* To set "not equals to," it is _not_ "<>" like you might think: it's `!=`.

## Other methods

There is also the function `which()` that you may see around, but in general I would recommend not using this since it relies on numeric subsetting and can be difficult to debug.

And, if you are regular expression junkie, you can use `grepl()` in your row or column selections (typically, it's in your row selection). There is a `grep()` function that actually returns the matched values, but, if you're doing a selection, you actually want to return a _logical vector_ (`TRUE`s and `FALSE`s) for your condition as to which rows you want to match...and that is what `grepl()` does.

If you have loaded `dplyr()` then it makes sense to use its `select()` for columns and `filter()` for rows instead.

## Munging data

So, now you can subset at will, how does this apply to data munging?

Well, in many cases your data will come with elements you need to change that you need to filter down to.  You can then reassign those values to what you prefer.

A few other functions are useful to know for these cases:

```{r}
## Will return TRUE if a value is NA (e.g. imported incorrectly)
is.na(NA)

a_vector <- c(1,2,3,NA,4)
is.na(a_vector)
```

## Munging Example

Lets take the previous mtcars columns and say we want to set all the mpg values to 24 if they are over 24.

In this case we can filter to the elements we need like before, but this time modifying the data in place using the `<-` assignment command:

```{r}
my_new_data <- mtcars
my_new_data[my_new_data$mpg > 24, "mpg"] <- 24
max(my_new_data[, "mpg"])
```
