---
title: "R Classes"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r classes-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We've touched on this subject already through the previous chapters, but a more detailed look at the type of objects R uses e.g. an object's `class` is useful.

# Useful class functions

When working with classes, its good to know these functions:

```r
## str() gives you the "Structure" of an object, including its class
str(mtcars)

## class() lets you query an objects class directly
class(mtcars)

```

# Atomic classes

There are many different types of `class`, and you can make your own, so this is not a definitive list, but will help in 95% of cases:

* `numeric` - a number like `1 2 3`
* `character` - a character string like `hello`
* `logical` - `TRUE` or `FALSE`
* `Date` - an object that represents date.  Note this is where it gets tricky, since when you print a `Date` object and a `character` with the same format, they look the same, but are different!

```{r}
a_date <- Sys.Date()
a_date
class(a_date)

a_character <- "2016-08-24"
a_character
class(a_character)

```

Things like finding date ranges or weekdays will work on a `Date` object, but not on a `character`

* `factor` - A categorical varaible.  

This is another tricky one, as they look the same as `character` when printed but act quite differently.

```{r}
a_factor <- factor("hello", levels = c("hello","goodbye"))
a_factor
class(a_factor)

a_string <- "hello"
a_string
class(a_string)
```

As an example, see what happens when we try to add a string and a factor together in the same vector:
```{r}
## an unexpected result??
c(a_factor, a_string)
```

Whats going on?  Since `a_factor` is a factor, it is actually represented as a number out of the choice of levels it could possibly be (`c("hello","goodbye")`).  When it is added to the character the factor is **coerced** into a character via `as.numeric`, and then into a character `as.character`.

Upshot of this all is to be very careful in making sure your variables are the class you expect them to be.

A classic mistake is to use `data.frame()` or `read.csv()` to make a data.frame from your data, but to not set the `stringsAsFactors = FALSE` argument, which if not used will default to using factors instead.

# Multi-classes

There are also objects in R that work with combinations of the classes above.

## Vector

A vector is a combination of the atomic elements above.  You can only combine elements of the same type in a vector.

You create a vector using `c()`

```{r}
a_vector <- c("a","b","c","d")
a_vector
class(a_vector)
str(a_vector)
```

> The class of the vector is the same as the element!  

This hints at a powerful feature of R, namely **vectorisation**.  The atomic elements above are actually a one-length vector.  This means that anything you can do to one element, will also work on an entire vector of the same `class`.

An example of this:

```{r}
## sum individual elements
sum(1,2,3,3,4,5,6)

## sum a vector
a_vector <- c(1,2,3,3,4,5,6)
sum(a_vector)

```

Some useful shortcuts with vectors are below:

```{r}
## make a sequence
1:10

## the lowercase letters
letters

## uppercase
LETTERS

```


## data.frame

The most common and useful R class is the `data.frame`

```{r}
class(mtcars)
str(mtcars)
```

This is most often used to represent tabular data, and very often used in R functions.  

You make a data.frame using the `data.frame()` function:

```{r}
## names before equals, values after
my_data_frame <- data.frame(numbers = 1:5, 
                            letters = c("a","b","c","d","e"), 
                            logic = c(TRUE, FALSE, FALSE, TRUE, TRUE))
class(my_data_frame)
str(my_data_frame)

```

> Each column can only be one class, but the class of the columns can be different types.

Also see that by default it will turn characters into `factors`.  Turn this off via the `stringsAsFactors = FALSE` argument:

```{r}
## names before equals, values after
my_data_frame <- data.frame(numbers = 1:5, 
                            letters = c("a","b","c","d","e"), 
                            logic = c(TRUE, FALSE, FALSE, TRUE, TRUE),
                            stringsAsFactors = FALSE)
class(my_data_frame)
str(my_data_frame)

```

You can access the individual columns of a `data.frame` via the `$` notation:

```{r}
## the column of numbers
my_data_frame$numbers
class(my_data_frame$numbers)
```

`data.frame`'s are a special case of the next object, `lists()` where all the columns are equal length.

## list

A list is like a data.frame, but can carry variable lengths of objects.  List elements can be anything, including `data.frames` or other `lists`

```{r}

my_list <- list(letters = letters,
                numbers = 1:5,
                all_data = my_data_frame,
                nested = list(LETTERS))
class(my_list)
str(my_list)
```

Like `data.frames` you can reach individual items via the `$` symbol:

```{r}
extract <- my_list$all_data
class(extract)
str(extract)
```

# Coercing

If you find an R object the wrong class for the function you need, what can you do?  This is where **corecian** comes into play.

All the class have an `as.this` function, which when you pass an R object in will try to change it to what you need.  It will usually error if this is impossible (which is much better than failing silently!)

Some coercing functions as shown below:

```{r}
## quotes indicate characters
as.character(-1:3)

## 0 is FALSE, everything else is TRUE
as.logical(-1:3)

## character to date
as.Date("2015-01-02")

## If your dates are in a different format to YYYY-MM-DD then you need to supply format argument
as.Date("20150102", format = "%Y%m%d")
as.Date("12-24-2016", format = "%m-%d-%Y")

## to change factors to numeric, be careful to go via as.character first
numeric_factor <- factor(1, levels = 5:1)
numeric_factor

## it gives answer as 5, as thats the first factor
wrong_factor <- as.numeric(numeric_factor)
wrong_factor

## we go via as.character, to get whats expected
right_factor <- as.numeric(as.character(numeric_factor))
right_factor

```
