---
title: "Correlation"
---

Correlation is a number that describes how much relationship is between two variables.

In digital analytis terms, you can use it to explore relationships between web metrics to see if an influence can be inferred, but beware the 101 stat advice of not accounting for other factors.  

For instance, a high correlation between social shares and SEO position could mean:

* Social shares influence SEO position
* SEO position influences social shares
* Social shares and SEO position are influenced by a third factor (such as Brand strength)
* The relationship was a chance error

It is rare to see the last three possibilties considered.

## Performing correlation analysis in R

That said, correlations can really help get an understanding of your data and hint at relationships.

The base function `cor()` will perform correlations on a `data.frame`.

The below example uses the web data dataset used in Excercise 4:

```{r}
## used for kable that makes nice RMarkdown tables
library(knitr)
web_data <- read.csv("./data/gadata_example_2.csv", stringsAsFactors = FALSE)
kable(head(web_data[-1]), row.names = FALSE)
```

Correlations will only work with numeric data, so we subset to just those columns, and then run the base R function `cor()` to see a correlation table:

```{r}
web_data_metrics <- web_data[c("sessions","pageviews","entrances","bounces")]
## see correlation between all metrics
kable(cor(web_data_metrics))
```

The table is mirrored in the diagonal. 1 means a perfect correlation, 0 no correlation and -1 a perfect negative correlation.

When working with correlations, its always a good idea to view an exploratory plot.  A handy function for this is `pairs()` which creates a scatter plot of all the metrics passed in combination:

```{r}
pairs(web_data_metrics)
```

Here you can see the correlation numbers in graphical form - for instance the high correlation of 0.9999923 between sessions and entrances results in an almost perfect straight line.  As a session starts with an entrance, this makes perfect sense.  A correlation of less than 1 may be a quick diagnostic that something is wrong with the tracking.

## How do web channels correlate?

One useful piece of analysis is seeing how web channels possibly interact.

### Data prep

To get the data in the right format, the below code pivots via the `reshape2` package:

```{r}
## use tidyverse to pivot the data
library(dplyr)
library(tidyr)

## get only desktop rows, and the date, channelGrouping and sessions columns
pivoted <- web_data %>% 
  filter(deviceCategory == "desktop") %>% 
  select(date, channelGrouping, sessions) %>%
  spread(channelGrouping, sessions)

## get rid of any NA's and replace with 0
pivoted[is.na(pivoted)] <- 0

kable(head(pivoted))
```

### Examining data

We can plot and correlate all the metrics for an overview:

```{r}
## not including first column, so -1 subset
cor_table <- round(cor(pivoted[-1]),2)
kable(cor_table)

pairs(pivoted[-1])
```

### Analysis

Now we compare channels we can see much looser correlations for this dataset, which is typical.  Correlations under 0.3 are as a rule-of-thumb not worth considering, so the standouts look to be Social vs Video and Paid vs Organic Search.

Plotting those channels we can examine the trends to see the shape of the data

> Correlation has help us zero in on possible interesting relationships


```{r}
library(ggplot2)
gg <- ggplot(data = pivoted) + theme_minimal() + ggtitle("Paid (blue) vs Organic (green) search")
gg <- gg + geom_line(aes(x = as.Date(date), y = `Paid Search`), col = "blue")
gg + geom_line(aes(x = as.Date(date), y = `Organic Search`), col = "green")
```

We can see here the trends do look similar, but with a paid search peak in the first Q.

```{r}
library(ggplot2)
gg <- ggplot(data = pivoted) + theme_minimal() + ggtitle("Social (red) vs Video (orange)")
gg <- gg + geom_line(aes(x = as.Date(date), y = Social), col = "red")
gg + geom_line(aes(x = as.Date(date), y = Video), col = "orange")
```

Here a peak in social late in the year looed to have conincided with a peak in social, possibly a campaign driving video views?

## Cross correlation

The correlations above all compare the same date point, but what if you expect a lagged effect?  Perhaps the video traffic drove social traffic later on due to client advocacy?

Cross correlations are useful when dealing with time-series data, and can examine if a metric has an influence on itself or another after some time has passed.  

This can be a powerful way to find if say a TV or Display campaign increased SEO traffic the next few weeks afterwards.

The below compares paid search on SEO - the correlation for different lags of days is the result - we can see a correlation at 0 lag at around 0.5, but the correlation increases if you lag the Social trend up to 10 days before. 

```{r}
ccf(pivoted$Social, pivoted$Video)
```

You could then conclude that Video was having a lagged effect on Social traffic up to 10 days beforehand, but beware - the nature of cross-correlation is that if both datasets have a similar looking spike, cross-correlation will highlight it.  Careful examination of the raw data trends should be done to verify it.  In some cases a smoother line will help get rid of spikes that effect the data (e.g. do the analysis on weekly or monthly data instead of daily.)
