---
title: "Day 2 in R"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Load the data

`readxl` is recommended as the best way to read in data.  You can also use the `Import Dataset` button in `RStudio` that will help you set options and write the `readxl` code for you:

```{r}
all_data <- readxl::read_excel("data/day2/SampleGA2 for R Mark.xlsx")

summary(all_data %>% 
          select(date, deviceCategory, channelGrouping, Sessions, GoalCompletions))
```
## First plot to find shape of the data

```{R}

overall_data <- all_data %>% 
  select(date, deviceCategory, channelGrouping, Sessions) %>% 
  mutate(yearMonth = format(date, format = "%Y%m")) %>% 
  group_by(yearMonth) %>% 
  summarise(sessionSum = sum(Sessions))

plot(overall_data, type = "l")

## a ggplot version
library(ggplot2)
gg <- ggplot(overall_data, aes(yearMonth, sessionSum)) + theme_dark()
gg <- gg + geom_line(aes(group = 1))
gg + geom_vline(xintercept=c(6,10), linetype = "dotted")
```

## Splitting into Groups

The statistical analysis was carried out on three subsets of this data. (see data two content)

The three files match to the following rows:

* Group 1 (row 1 - row 4054)
* Group 2 (row 4055 - 5905)
* Group 3 (row 5906 - 7162)

We want to run analysis on these groups seperately, so we split them up into three different data frames:

```{r}
## 1 indexed not including header row (1)
group_one <- all_data[1:4054, ]
group_two <- all_data[4055:5905, ]
group_three <- all_data[5906:7162, ]
```

Because we want to run analysis at the same time, we put those data.frames into a list.  This means we can run functions over them using `lapply`

```{r}
all_groups <- list(group_one = group_one,
                   group_two = group_two,
                   group_three = group_three)
```

## Cross Tabulation


```{r}
my_cross_table <- function(x, cross1, cross2){
  totals <- table(x[[cross1]], x[[cross2]])
  
  totals
}

lapply(all_groups, my_cross_table, cross1 = "deviceCategory", cross2 = "channelGrouping")
```


## Summary statistics

We start with creating summary statistics.  One option would be to use `summary()` but `describe()` from the `psych` package as it also shows `skew` and `kurtosis`.

The output is generated via the same general workflow:

1. Create a function that will operate on one of the group data.frames (e.g `all_groups[[1]]`)
2. Run the function over the list of group data.frames
3. Analyse the result

```{R}
library(psych) # statistics

my_summary <- function(x, metric){
  y <- x[, metric]

  describe(y)
}


lapply(all_groups, my_summary, metric = c("Sessions", "GoalCompletions"))

```

Note that we pass the `metric` argument via `lapply()`, which is constant for all the data passed to the functions.

### Create class statistics

To create other statistics, we need to create a new function:

```{r}
my_class_stats <- function(x, dimension){
  dim <- x[[dimension]]
  
  ## makes a dataframe of $Var1 and $Freq
  my_df <- as.data.frame(table(dim))
  sum_all <- sum(my_df$Freq)
  my_df$percent <- round(my_df$Freq / sum_all, 4)
  my_df$cum.percent <- cumsum(my_df$percent)
  
  my_df
}
```

To run the function over the different dimensions, we change the function argument:

```{r}
lapply(all_groups, my_class_stats, dimension = "deviceCategory")
lapply(all_groups, my_class_stats, dimension = "channelGrouping")
```



## Chi Squared and expected value

Here we first print out the expected values (Compare with above) and then perform a chi-squared test on if the actual values are signficantly different.

```{R, message=FALSE, warning=FALSE}
library(questionr) ## for cramerV

my_chi_table <- function(x, cross1, cross2){
  tab <- table(x[[cross1]], x[[cross2]])
  y <- chisq.test(tab)
  
  list(
    expected = y$expected,
    observed = y$observed,
    difference = y$expected - y$observed,
    chi = y,
    cramerV = round(cramer.v(tab), digits = 4)
  )

}

lapply(all_groups, my_chi_table, cross1 = "deviceCategory", cross2 = "channelGrouping")
```

### A nice plot 

```{r}
## get the chi-squared results into its own list
plot_this_list <- lapply(all_groups, my_chi_table, cross1 = "deviceCategory", cross2 = "channelGrouping")

plot_func <- function(x){
  barplot(x$difference, 
          beside = TRUE, 
          main = "Expected - Observed", 
          sub = paste(x$chi$method, x$chi$p.value, " CramerV: ", x$cramerV),
          col = c("#91210E","#A9E317","#2F5CAA"))
  legend("topright",fill=c("#91210E","#A9E317","#2F5CAA"), legend=rownames(x$difference))
}

lapply(plot_this_list, plot_func)

```

## ANOVA

Analysis of variance applied to the groups

```{r}

my_anova <- function(x, metric, dim){
  my_form <- as.formula(paste(metric, "~", dim))
  model <- aov(my_form , data = x)
  summary(model)
}

my_anova_tukey <- function(x, metric, dim){
  my_form <- as.formula(paste(metric, "~", dim))
  model <- aov(my_form , data = x)
  
  TukeyHSD(model)
}

my_nway_anova <- function(x, metric, dim){
  my_form <- as.formula(paste(metric, "~", dim))
  anova(lm(my_form , data = x))
}

lapply(all_groups, my_anova, metric = "Sessions", dim = "deviceCategory")
lapply(all_groups, my_anova, metric = "GoalCompletions", dim = "deviceCategory")


lapply(all_groups, my_anova, metric = "Sessions", dim = "channelGrouping")
lapply(all_groups, my_anova, metric = "GoalCompletions", dim = "channelGrouping")


## tests of between subject effects
lapply(all_groups, my_nway_anova, metric = "GoalCompletions", dim = "channelGrouping * deviceCategory")


```


## Regression outputs

```{r}
my_models_fun <- function(x, dependent, independent){
  my_form <- as.formula(paste(dependent, "~", independent))
  lm(my_form, data = x)
}

my_models <- lapply(all_groups, my_models_fun, dependent = "GoalCompletions", independent = "Sessions")
lapply(my_models, summary)

my_predict_mean <- function(x, model_string = "GoalCompletions ~ Sessions", new_value = mean(x$Sessions)){
  my_form <- as.formula(paste(model_string))
  model <- lm(my_form, data = x)
  cat("\nMean Sessions: ",mean(x$Sessions))
  cat("\nMean GoalCompletions",mean(x$GoalCompletions))
  predict(model, 
          newdata = data.frame(Sessions = mean(x$Sessions)),
          interval = "predict",
          level = 0.95)
}

my_conf_interval <- function(x){
  model <- lm(Sessions ~ GoalCompletions + DeviceClass + ChannelClass, data = x)
  summary(model)
  confint(model, level = 0.95)
  
}

my_regression <- function(x){
  model <- lm(Sessions ~ GoalCompletions + DeviceClass + ChannelClass, data = x)
  summary(model)
}



## generate the upper and lower bound for beta coeffcients
lapply(all_groups, my_predict_mean)

lapply(all_groups, my_regression)
lapply(all_groups, my_conf_interval)
```