---
title: "Example 1: Trending Google Analytics Sessions"
---

```{r example1-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a pretty straightforward example. It simply walks through getting daily sessions data from Google Analytics, and then a few different ways to trend it. Is it a sexy example? No. Or yes. You be the judge.

We first load the libraries we will need later, authenticate with Google, and set the View Id we want to pull data from. 

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(googleAnalyticsR)
library(knitr)

cache_file <- "./data/example1_gadata.RData"

if(file.exists(cache_file)){
  ## avoid calling API
  load(cache_file)
} else {
  ## call API, save data in cache
  ga_auth()

  ## replace with your own Google Analytics View Id
  view_id <- 81416156

  trend_data <- google_analytics_4(view_id,
                                 date_range = c(Sys.Date() - 400, Sys.Date()),
                                 dimensions = c("date"),
                                 metrics = "sessions",
                                 pivots = pivot_ga4("medium","sessions"),
                                 max = -1)
  save(trend_data, file = "./data/example1_gadata.RData")
  
}


```

```r
library(googleAnalyticsR)
library(knitr)
ga_auth()

## replace with your own Google Analytics View Id
view_id <- 81416156

```

This is the actual API call.  We take advantage of Google Analytics v4's pivot function, to do some data transformation for us:

```r
## get data, use pivot to get columns of medium sessions
trend_data <- google_analytics_4(view_id,
                                 date_range = c(Sys.Date() - 400, Sys.Date()),
                                 dimensions = c("date"),
                                 metrics = "sessions",
                                 pivots = pivot_ga4("medium","sessions"),
                                 max = -1)

## Note the medium columns created by the pivot
head(trend_data)
```

```{r, echo=FALSE}
kable(head(trend_data))
```

Now we have the data, we transform it a little to a format suitable for the plot functions:

```{r}
## format the names
names(trend_data)
names(trend_data) <- c("Date","Total","Referral","Direct","Social")

## check the data R classes
## This time all look good - Date objects for date and numeric for the metrics
str(trend_data)
```

Looks good to try some plots:

### Base plots

Good for quick explorations, but takes arcane work to make presentable for final publication

```{r}
## make a time-series object
trend_ts <- ts(trend_data[-1], frequency = 7)
plot(Total ~ Date, data = trend_data, type = "l")
axis(1, trend_data$Date, format(trend_data$Date, "%b"))
```

### ggplot

You need to learn the syntax, but its more consistent than base R's.  
You slowly build up your plot via geoms, and the end results can be very professional looking.

```{r}
library(ggplot2)
library(reshape2)

## change data into 'long' format
trend_long <- melt(trend_data, id = "Date")
head(trend_long)

## build up the plot
gg <- ggplot(trend_long, aes(x = Date, y = value, group = variable)) + theme_minimal()
gg <- gg + geom_line(aes(colour = variable))
gg
```

### highcharter (example of htmlwidgets)

Uses JavaScript libraries to render your data.  Only works in HTML, but fancy and interactive. 

```{r, message=FALSE, warning=FALSE}
library(highcharter)
library(reshape2)

## change data into 'long' format
trend_long <- melt(trend_data, id = "Date")
hchart(trend_long, type = "line", x = Date, y = value, group = variable)

```
