---
title: "Example 5: Visualizing Sessions in a Heatmap"
---

```{r example1-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

This example shows a visualization of traffic to a site broken down by day of week and hour of day.

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(googleAnalyticsR)
library(dplyr)
library(tidyr)
library(highcharter)


## If a cache file exists, load it from there
if(file.exists("./data/index_gadata.RData")){
  load("./data/index_gadata.RData")
} else {
  
  ## Otherwise, fetch  the data using the Google Analytics API.
  view_id <- 81416156
  ga_auth()

  gadata <- google_analytics_4(view_id, 
                             date_range = c(Sys.Date() - 400, Sys.Date()),
                             metrics = "sessions", 
                             dimensions = c("date","hour"),
                             max = -1)

  gadata$weekday <- ordered(weekdays(gadata$date, FALSE), 
                          levels = c("Monday","Tuesday","Wednesday",
                                     "Thursday","Friday","Saturday","Sunday")) 
  ## save cache
  save(gadata, file = "./data/index_gadata.RData")
  
}

```

```r
# Start by loading the libraries we'll want to use.
library(googleAnalyticsR)
library(dplyr)
library(tidyr)
library(highcharter)

# Set the view ID that we'll be using. You can get the view ID for a specific view
# that you have access to by logging into the Google Analytics Query Explorer at
# https://ga-dev-tools.appspot.com/query-explorer/. It's the "ids" value.
view_id <- 81416156

# Authorize Google Analytics
ga_auth()

# Pull the data. This is set to pull the last 400 days of data.
gadata <- google_analytics_4(view_id, 
                             date_range = c(Sys.Date() - 400, Sys.Date()),
                             metrics = "sessions", 
                             dimensions = c("date","hour"),
                             max = -1)

# Added a column to the data with the weekday.
gadata$weekday <- ordered(weekdays(gadata$date, FALSE), 
                          levels = c("Monday","Tuesday","Wednesday",
                                     "Thursday","Friday","Saturday","Sunday")) 
```
Let's look at the first few rows (the "head") of our data to see where we stand so far:

```{r initial-data, echo=FALSE}
kable(head(gadata))
```

Now that we have the data, we just have to do a little bit of additional work to get the data arranged properly to generate the heatmap. The code for this could be a lot simpler/shorter, but it's split out here into separate lines to make it easier to follow. Trust us -- this looks like it's a lot, but it's really just doing the equivalent of generating a pivot table with a very specific structure:

```{r, message=FALSE, warning=FALSE}

# Subset the data to just be the weekday, hour of the day, and sessions. (This
# means we're getting rid of the "date" column)
heatmap_data <- select(gadata, weekday, hour, sessions)

# Summarize the data by weekday-hour
heatmap_sums <- group_by(heatmap_data, weekday, hour) %>%
  summarise(sessions = sum(sessions))

# Now, "spread" the data out so it's heatmap-ready
heatmap_recast <- spread(heatmap_sums, hour, sessions)

# Make this "data frame" into a "matrix"
heatmap_matrix <- as.matrix(heatmap_recast[-1])

# Name the rows to match the weeks
row.names(heatmap_matrix) <- c("Monday","Tuesday","Wednesday",
                               "Thursday","Friday","Saturday","Sunday")

# Generate the heatmap of weekdays per hour
hchart(heatmap_matrix, type = "heatmap")

```

There we have it! Mouse over it to see the specific values.