---
title: "Exercise 5 -- Correlation"
output:
  html_document:
    toc: true
    toc_float: true

---

This exercise is dependent on the successful completion of [Exercise 4](exercise4-tidy-dplyr.html).

For this exercise, complete the following tasks with `web_data` (which should now have a `pageviews_per_visit` and a `bounce_rate` column following Exercise 4):

1. Evaluate the correlation between all of the metrics `web_data`.
2. Evaluate the correlation between pageviews for each `deviceCategory`.

-------------

```{r echo=FALSE, message=FALSE}
# We'll be using 'dplyr' functions
library(dplyr)

# Replicate the results from Exercise 2 so we've got the data
# We need to work with.
web_data <- read.csv("data/gadata_example_2.csv", stringsAsFactors = FALSE) %>%
  mutate(pvs_per_session = pageviews / sessions, 
                  bounce_rate = bounces / entrances)
```

## 1. Bounce Rate vs. Sessions
Remember, this exercise is contingent on having completed [Exercise 4](exercise4-tidy-dplyr.html). Check your **Environment** tab to make sure you have a `web_data` data frame with 5732 rows (observations) and 10 variables. Or, use the `str()` function to check the structure of the object:

```{r}
str(web_data)
```

All set? Good.

```{r message=FALSE}
# We'll be using 'dplyr' functions
library(dplyr)

# Get just the data -- numeric only! -- that we'll be working with
cor_data <- select(web_data, -c(X, date, deviceCategory, channelGrouping))

# Get a table of correlation coefficients.
cor(cor_data)
```

Let's go ahead and get a visual of the scatterplots of this:

```{r}
pairs(cor_data)
```

There are several things we could dig into there (for instance: what's going on with pageviews and sessions? It looks like two distinct patterns there that might be worth exploring), but...the assignment is the assignment, so let's move on!

## 2. Pageviews for Device Type
From 

```{r message=FALSE}

library(dplyr)
library(tidyr)

# Collapse the channelGrouping by summing them, and then "spread" the 
# deviceCategory values over the columns
pivoted_data <- 
  summarise(group_by(web_data, date, deviceCategory), 
            pageviews = sum(pageviews)) %>%
  spread(deviceCategory, pageviews)

# Get rid of any NA's and replace with 0
pivoted_data[is.na(pivoted_data)] <- 0

# Let's take a quick look at our data
head(pivoted_data)
```

Now, we can calculate the correlation coefficients and make some scatterplots!

```{r}
# Remember, we don't want the date field, so we have to remove that
cor(pivoted_data[,-1])
pairs(pivoted_data[,-1])
```

The interpretation of this isn't particularly sexy, as it makes sense that, when traffic increases for one device type, it probably increases for other device types. But, since the pedagocial purpose of this exercise was to get you subsetting the data and digging in, we're going to call it a success!