---
title: "Exercise 6 -- Anomaly Detection"
output:
  html_document:
    toc: true
    toc_float: true

---

This exercise is dependent on the successful completion of [Exercise 2](exercise2-io.html).

For this exercise, use the `AnomalyDetection` package to identify anomalies in the `web_data` set -- looking for anomalies in each combination of `deviceCategory`-`channelGrouping`:

1. Write a script to look for anomalies in the `sessions` data.
2. Modify the script to look for anomalies in the `bounceRate`.

Sound intimidating? It's actually fairly simple!

-------------

```{r echo=FALSE, message=FALSE}

# Replicate the results from Exercise 2 so we've got the data
# We need to work with.
web_data <- read.csv("data/gadata_example_2.csv", stringsAsFactors = FALSE) 
```

## 1. Anomalies in Sessions

First, we need to install and load the `AnomalyDetection` package, so type `install.packages("AnomalyDetection")` and, once it's installed, add this line to your script:

The easiest way to do this is to simply build a nested loop to run through each combination of `deviceCategory` and `channelGrouping`. Now, generally, it's better to take advantage of **vectorization** in R and _not use loops at all_...but that's just a hint as to which of your instructors wrote this exercise.

We can do this by getting the unique values of each using the `unique()` function. For now, let's write that code and then just demonstrate that these loops can crank through all the possibilities.

```{r message=FALSE, eval=FALSE}
# Get vectors with the unique values for each variable
device_cats <- unique(web_data$deviceCategory)
channel_grps <- unique(web_data$channelGrouping)

# Start with a loop for each unique value of deviceCategory
for (d in 1:length(device_cats)){
  
  # Then, add a loop within it for each unique channelGrouping
  for (c in 1:length(channel_grps)){
    
    # For now, let's just print all the values this loop creates.
    # the \n is just a new line indicator
    cat("deviceCategory =",
        device_cats[d],
        "and channelGrouping =",
        channel_grps[c],
        "\n",sep=" ")
  }
}
```

If you run that, you should see a list of all combinations of `deviceCategory` and `channelGrouping`

Now, let's right a separate bit of code just to prove to ourself that we can pull a subset of data that is _just_ the sessions for a given pair of values for these two variables. You can put this in place of the `cat()` function and just **Run** these new lines to try it out. Ultimately, we'll swap these rows out for the dynamic loop values:

```{r message=FALSE}
# We'll need dplyr
library(dplyr)

# Filter the data to only include a specified deviceCategory and 
# channelGrouping (we'll make these dynamic in a bit) and pull just
# the date, deviceCategory, channelGrouping, and sessions
current_data <- filter(web_data,
                       deviceCategory=="mobile",
                       channelGrouping=="Email") %>%
  select(date, deviceCategory, channelGrouping, sessions)

head(current_data)

```

So, good. That works! So, now, let's do three things:

1. Add the `AnomalyDetection` package
2. Run the `$sessions` data through the `AnomalyDetectionTs()` (time-series) function.

We just need to make a couple of updates to the code:

```{r message=FALSE}
# We'll need dplyr and AnomalyDetection
library(dplyr)
library(AnomalyDetection)

# Filter the data to only include a specified deviceCategory and 
# channelGrouping, and convert the date to POSIXct
current_data <- filter(web_data,
                       deviceCategory=="mobile",
                       channelGrouping=="Email") %>%
  select(date, sessions) %>%
  mutate(date2 = as.POSIXct(date)) %>%
  select(date2, sessions)

# Run anomaly detection on the data
AnomalyDetectionTs(current_data, direction = "both", plot=TRUE)

```

