---
title: "Welcome to DARTISTICS!"
author: "Digital Analytics: R and staTISTICS"
---

```{r index, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

This site is intended to be a resource for digital analysts who are interested in learning or expanding their knowledge of R and statistics. That's from whence the semi-silly name came from:

<div style= "background-color: #cce6ff; text-align: center; font-size: 20px; padding: 10px 10px 10px 10px;"><b>D</b>igital <b>A</b>nalysts: <b>R</b> and sta<b>TISTICS</b></div>

<br>_(Clearly, this is not a site about "branding an initiative" or "coming up with a totally awesome name for a site!")_

This site is developed and maintained by [Mark Edmondson](https://dk.linkedin.com/in/markpeteredmondson) and [Tim Wilson](https://www.linkedin.com/in/tgwilson), with semi-intentional assists from [Donal Phipps](https://uk.linkedin.com/in/donalp), [Dr. Michael Levin](https://www.linkedin.com/in/michael-levin-30479086/) and others. If you have feedback or suggestions for the site, [hop over to GitHub and log an issue](https://github.com/MarkEdmondson1234/dartistics.com/issues). _(A GitHub account is required, but, if you're using R, you should have one! And, creating one is quick and free if you don't.)_

Our hope is that, for the analyst who is new to R, by going through the resources on this site, the code and output below will start to make some real sense!

## Some Examples for Inspiration

Below are a couple of examples of some uses of R with web analytics data. Check out the examples under the **Examples** dropdown for these examples and more, including the code used to generate them.

### Visualizing Sessions in a Heatmap

This example shows a visualization of traffic to a site broken down by day of week and hour of day.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(highcharter)
library(googleAnalyticsR)
library(reshape2)
library(forecast)

## If a cache file exists, load it from there
if(file.exists("./data/index_gadata.RData")){
  load("./data/index_gadata.RData")
} else {
  
  ## Otherwise, fetch  the data using the Google Analytics API.
  view_id <- 81416156
  ga_auth()

  gadata <- google_analytics_4(view_id, 
                             date_range = c(Sys.Date() - 400, Sys.Date()),
                             metrics = "sessions", 
                             dimensions = c("date","hour"),
                             max = -1)

  gadata$weekday <- ordered(weekdays(gadata$date, FALSE), 
                          levels = c("Monday","Tuesday","Wednesday",
                                     "Thursday","Friday","Saturday","Sunday")) 
  ## save cache
  save(gadata, file = "./data/index_gadata.RData")
  
}

heatmap_data <- gadata[, c("weekday","hour","sessions")]

heatmap_recast <- dcast(heatmap_data, weekday ~ hour, sum)
heatmap_recastv <- dcast(heatmap_data, hour ~ weekday, sum)

heatmap_matrix <- as.matrix(heatmap_recast[-1])
heatmap_matrixv <- as.matrix(heatmap_recastv[-1])

row.names(heatmap_matrix) <- c("Monday","Tuesday","Wednesday",
                               "Thursday","Friday","Saturday","Sunday")
colnames(heatmap_matrixv) <- c("Mo","Tu","We","Th","Fr","Sa","Su")
```

<div class="heatwide">
```{r, message=FALSE, warning=FALSE, echo=FALSE}
## heatmap of weekdays per hour
hchart(heatmap_matrix, type = "heatmap")
```
</div>

<div class="heatnarr">
```{r, message=FALSE, warning=FALSE, echo=FALSE}
## heatmap of weekdays per hour
hchart(heatmap_matrixv, type = "heatmap")
```
</div>

### Forecasting with Holt-Winters
This example uses the Holt-Winters method to apply some smoothing and seasonality to the base data to build a forecast that includes the likely range of values.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
## if file exists, fetch data from there
if(file.exists("./data/index_gadata2.RData")){
  
  load("./data/index_gadata2.RData")
  
} else {
  ## fetch from API
  gadata2 <- google_analytics_4(view_id, 
                             date_range = c("2013-08-01", "2016-07-31"),
                             metrics = "sessions", 
                             dimensions = c("yearMonth"),
                             max = -1)
  save(gadata, file = "./data/index_gadata2.RData")
  
}

ga_ts <- ts(gadata2$sessions, start = c(2013,08), end = c(2016,07), frequency = 12)

forecast1 <- HoltWinters(ga_ts)

## forecast for next 12 months of the blog sessions
hchart(forecast(forecast1, h = 12))

```

These are just two of an infinite number of examples of what can be done with the power of R, a bit of knowledge of statistical concepts, and access to web analytics data. Happy coding!

