---
title: "Welcome to DARTISTICS!"
author: "Digital Analytics: R and staTISTICS"
---

```{r index, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This site is intended to be a resource for digital analysts who are interesting to learn or expand their knowledge of R and statistics. That's from whence the semi-silly name came from:

<div style= "background-color: #cce6ff; text-align: center; font-size: 20px; padding: 10px 10px 10px 10px;"><b>D</b>igital <b>A</b>nalysts: <b>R</b> and sta<b>TISTICS</b></div>

_(Clearly, this is not a site about "branding an initiative" or "coming up with a totally awesome name for a site!")_

This site is developed and maintained by [Mark Edmondson](https://dk.linkedin.com/in/markpeteredmondson) and [Tim Wilson](https://www.linkedin.com/in/tgwilson), with a semi-intentional assist from [Donal Phipps](https://uk.linkedin.com/in/donalp). If you have feedback or suggestions for the site, [hop over to GitHub and log an issue](https://github.com/MarkEdmondson1234/dartistics.com/issues). _(A GitHub account is required, but, if you're using R, you should have one! And, creating one is quick and free if you don't.)_

Our hope is that, for the analyst who is new to R, by going through the resources on this site, the code and output below will start to make some real sense!

### 3-Day Training: Columbus, Ohio, US - June 13-15, 2017

<div style= "background-color: #cce6ff; text-align: center; padding: 10px 10px 10px 10px;"><b>See the <a href="r-and-statistics-class.html">class details</a> for more information.</b></div><br>

### Visualizing Sessions in a Heatmap

This example shows a visualization of traffic to a site broken down by day of week and hour of day.

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(highcharter)
library(googleAnalyticsR)
library(reshape2)
library(forecast)

## If a cache file exists, load it from there
if(file.exists("./data/index_gadata.RData")){
  load("./data/index_gadata.RData")
} else {
  
  ## Otherwise, fetch from API.
  view_id <- 81416156
  ga_auth()

  gadata <- google_analytics_4(view_id, 
                             date_range = c(Sys.Date() - 400, Sys.Date()),
                             metrics = "sessions", 
                             dimensions = c("date","hour"),
                             max = -1)

  gadata$weekday <- ordered(weekdays(gadata$date, FALSE), 
                          levels = c("Monday","Tuesday","Wednesday",
                                     "Thursday","Friday","Saturday","Sunday")) 
  ## save cache
  save(gadata, file = "./data/index_gadata.RData")
  
}

```

```r
library(highcharter)
library(googleAnalyticsR)
library(reshape2)
library(forecast)

view_id <- 81416156
ga_auth()

gadata <- google_analytics_4(view_id, 
                             date_range = c(Sys.Date() - 400, Sys.Date()),
                             metrics = "sessions", 
                             dimensions = c("date","hour"),
                             max = -1)

gadata$weekday <- ordered(weekdays(gadata$date, FALSE), 
                          levels = c("Monday","Tuesday","Wednesday",
                                     "Thursday","Friday","Saturday","Sunday")) 
```

```{r, message=FALSE, warning=FALSE}

heatmap_data <- gadata[, c("weekday","hour","sessions")]

heatmap_recast <- dcast(heatmap_data, weekday ~ hour, sum)

heatmap_matrix <- as.matrix(heatmap_recast[-1])

row.names(heatmap_matrix) <- c("Monday","Tuesday","Wednesday",
                               "Thursday","Friday","Saturday","Sunday")

## heatmap of weekdays per hour
hchart(heatmap_matrix, type = "heatmap")

```

```{r, message=FALSE, warning=FALSE, include=FALSE}
## if file exists, fetch data from there
if(file.exists("./data/index_gadata2.RData")){
  
  load("./data/index_gadata2.RData")
  
} else {
  ## fetch from API
  gadata2 <- google_analytics_4(view_id, 
                             date_range = c("2013-08-01", "2016-07-31"),
                             metrics = "sessions", 
                             dimensions = c("yearMonth"),
                             max = -1)
  save(gadata, file = "./data/index_gadata2.RData")
  
}

```

### Forecasting with Holt-Winters
Long gone (can) be the days of forecasting simply by dropping a trendline on some data. This example uses the Holt-Winters method to apply some smoothing and seasonality to the base data to build a forecast that includes the likely range of values.

```r
gadata2 <- google_analytics_4(view_id, 
                             date_range = c("2013-08-01", "2016-07-31"),
                             metrics = "sessions", 
                             dimensions = c("yearMonth"),
                             max = -1)
```

```{r, message=FALSE, warning=FALSE}
ga_ts <- ts(gadata2$sessions, start = c(2013,08), end = c(2016,07), frequency = 12)

forecast1 <- HoltWinters(ga_ts)

## forecast for next 12 months of the blog sessions
hchart(forecast(forecast1, h = 12))

```


