---
title: "Intro to R"
author: "DonalPhilpps"
date: "28 June 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

#  Basics

## Creating variables

```{r}
x<-"hello, world!"
x
print(x)
```


## Working directory

Dangers of setting setwd() in scripts on interportability, perhaps explanation on using RStudio projects

```{r}
getwd()
# setwd()
list.files()
``` 

## Data types

### Numeric data types  

```{r}
test<-1
test2<-c(1,2,3,4,5,6,7)
mean(test2)
median(test2)
test3<-c(1:7)
test2==test3
test2*test3
```

### Character class

```{r}
x<-c("hello","world")
class(x)
x[1]
paste(x[1], x[2], "from London.")
```

### Booleans / logicals

```{r}
bool<-c(TRUE,FALSE,FALSE,TRUE,FALSE,FALSE,TRUE)
bool
```

### Matrix

```{r}
mat<-matrix(1:100, ncol=10, nrow=10)
mat

mat[ 1, ]
mat[ , 1 ]
```

### Data frames

One of the most important R objects, this standardises data for use in packages

```{r}
# Recall our test2 vector of numbers, and our bool vector of Booleans
test2
bool
test2[bool]

comb<-data.frame(test2,bool)
mat.frame<-data.frame(mat)
mat.frame$X1
mat.frame$X4
```      

### combining data types, not possible in matrices

```{r}
mat.frame<-cbind(mat.frame,c("one","two","three","four","five","six","seven","eight","nine","ten"))
names(mat.frame)<-c("ones","tens","twenties","thirties","fourties","fifties","sixties","seventies","eighties","nineties", "words")
#accessing dataframe columns by colnames
mat.frame$twenties
mat.frame[1:3,"thirties"]
# simple arithmetic
mat.frame$tens * mat.frame$twenties
```


### Lots of data

```{r}
biggie1<-rnorm(100000, 5, 5)
biggie2<-rnorm(100000, 5, 100)
mean(biggie1)
big.frame<-data.frame(cbind(biggie1,biggie2))
ggplot(big.frame,aes(x=biggie1,y=biggie2)) + geom_point(alpha=0.04)

```

# Reading data



```{r}
# reads in some example data from the GA API
GAreport<-read.csv("data/gadata_example.csv", stringsAsFactors=FALSE) 
head(GAreport)

# Removing unwanted columns
GAreport2 <- GAreport[, -(1:2)]

# Simple data examinations made easy
summary(GAreport)
quantile(GAreport$sessions)
str(GAreport)
```

## Quick exploratory plots

```{r}
plot(GAreport$sessions, GAreport$ratio)

hist(GAreport$sessions, main="Histogram of all sessions", xlab="Sessions")
abline(v=mean(GAreport$sessions), col="blue")
abline(v=median(GAreport$sessions),col="red")
```

## More professional plots

```{r}
# nicer reports
library(ggplot2)

## I don't like using qplot as its a bit of crutch when you need to customise later
# qplot(GAreport$sessions, main="Histogram of all sessions", xlab="Sessions")

ggplot(GAreport, aes(x=sessions)) + geom_histogram(binwidth = 10, fill = "purple") + ggtitle("Histogram of all sessions")

```


# Glimpse at improved statistical power

*first - refresher on data types (exaplained verbally)*
We need x axis to be a date and not a category

```{r}
## date is currently a character as imported from a csv
str(GAreport)

g<-ggplot(GAreport, aes(x=as.Date(date), y=sessions)) + geom_line() + ggtitle("Example sessions by date")
g
g + geom_smooth()

# Super-simple regression model - do sessions increase over time?
quickModel<-lm(sessions ~ as.Date(date), GAreport)
summary(quickModel)
# Plotting a linear model
g + geom_smooth(method = "lm", se = FALSE)
```  
