---
title: "Learn to love lists"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lists

Lists are a key data type in R but can be avoided by beginners due to being a bit harder to work with.  The do however hold the key to many of R's more advanced features.

## Navigating lists

### [ ] vs [[ ]]

A key learning is to know the difference between `[ ]` and `[[ ]]` when subsetting lists.

When using `[ ]` you will always get out another list.  When using `[[ ]]` you will get out the content of that list.

```{r}

a_list <- list(head(mtcars), letters, list(a=1, b=2))
str(a_list)

a_list[1]
class(a_list[1])

a_list[[1]]
class(a_list[[1]])


```

This helps you make sure you are working on what you expect when dealing with list elements.

### Named lists

Lists can also have names, and it can be useful do to so you can call specific elements by name, using the `list$name` convention, which is equivalent to `list[["name"]]`

```{r}

a_list <- list(df = head(mtcars), abc = letters, b_list = list(a=1, b=2))
str(a_list)

a_list["df"]
class(a_list["df"])

a_list[["df"]]
class(a_list[["df"]])

```

#### Exercise

What is the code to extract the other elements of `a_list` ?

```r
a_list <- list(df = head(mtcars), abc = letters, b_list = list(a=1, b=2))


a_list_letters <- ## your code here

a_list_b <- ## your code here

## These shoud be TRUE
class(a_list_letters) == "character"
class(b_list) == "list"
```

### Viewing lists

#### str(x, max.level = , list.len = )

`str()` is your friend as usual, but when printing out lists can turn into console spam.  You can limit this by adding these parameters to `str()` to let you have more control on what is printed out. 

#### library(listviewer)

Another handy library to help with lists is `listviewer::jsonedit(your_list)` which gives you a nested structure to look through the list. 

```{r}
library(listviewer)
jsonedit(mtcars)
```

## The "apply" family functions

We move now to apply functions.  We will look at the three most popular: `lapply()`, `vapply()` and `apply()`

### lapply()

List apply will take each element of your list, call a function upon it, then return a list of the results.  `lapply()` can be really useful to write compact, understandable code.

It is often used to replace what would be `for` loops in other languages, and offers benefits like better memory management and vectorisation.  Its one reason some cite use of `for()` loops in R code as bad R-style, although there is still some scope for them.

#### Guide to using lapply()

Lets build up an lapply function, that will work on a list of `data.frame`'s.

1. Choose what you want the `lapply()` function to iterate across.  In this example we have a list of data.frames, `a_list_of_df`

2. Choose our function.  This will operate on each element,

3. Test the function on one element i.e. the contents of `a_list_of_df[[x]]` *NOT*  `a_list_of_df[x]`.  Once happy we know it should work on the entire list.

4. Any constant arguments get added to the end of the `lapply` - e.g. is you are using `sum()` you may want all the `sum` functions to ignore `NA` e.g. `sum(x, na.rm = TRUE)`

5. Construct the full function `result <- lapply(a_list, myfunction, constant_arguments = TRUE)`


#### Example

Here we would like to just check the number of rows for each data.frame, so the function chosen is `nrow()`.

Testing the function chosen against our example, we see it works as expected:

```r
nrow(a_list_of_df[[1]])
# 345
```

We can now apply this to all the data.frames in the list:

```r
lapply(a_list_of_df, nrow)
```

Note we don't add arguments to the function call (`nrow`, not `nrow()`).  

If we want to add arguments to the function, we can supply those as named arguments to `lapply()`

Here we take advantage that a data.frame is a list of equal length vectors:

```r
lapply(mtcars, mean, trim = 0.5)
```

#### Using your own function

What if you want to supply your own function?  You can predefine it before the lapply() and supply it:

```r
my_func <- function(x){
  sum(x)
}

lapply(my_vector, my_func)

```

...or you can add it straight into the lapply function itself - this is normally done when the function is small:

A function defined this way is called an `anonymous function`, as it has no name.

```r
lapply(my_vector, function(x){
  sum(x)
})

```

### vapply()

`vapply()` is an extension on top of `lapply()`.  It works in the same way, but you also supply a `FUN.VALUE` which is a template of what you expect the result to be.  If the result is not of that class, then it will raise an error.  

This is good to stop any nasty surprises messing up your code later, and is recommended over `lapply()` when creating functions.

It also has the `USE.NAMES` argument which is useful to have the list output have the same names as the input list. 

### apply()

`apply()` is similar but lets you work with data.frame rows as inputs, instead of list elements.  Here each element is a data.frame row, which you can then operate on from your supplied function:

```{r}
apply(mtcars, MARGIN = 2, sum)

```


### Exercise

Write a function that iterates over this list of `data.frames` supplied below, to output the `max` of each column.  The output should be a list of numeric vectors, each vector element being the maxium number.  

```r
## don't touch this, it just creates your data
a_list_of_df <- lapply(1:10, function(x){
    data.frame(matrix(runif(1000), ncol = 10))
})

## perhaps you want to pass more arguments in, replace as you see fit
your_function <- function(a_data_frame){
  
}

your_result <- lapply(a_list_of_df, your_function)

```

> Hint - use `colSums`

How would you change this to use `vapply` ?


## Working with the result of lists

There are some handy functions to work with lists: I use mostly `Reduce()`.  This takes a list in its second argument and applies the function given in its first argument in turn, adding the result of the previous call to the next. 

As an example, you can use `rbind()` which binds a data.frame to another and apply it to a list of data.frames.  Reduce will `rbind()` the first two data.frames, then take that result and `rbind()` it to the third, then take that result and `rbind()` it to the forth, etc.  until you have added up all the data.frames into one:

```r
Reduce(rbind, list_of_df)
```

### purrr() 

For more advanced useage, [`purrr`](https://github.com/tidyverse/purrr) is a tidyverse package which is very useful for both working with lists and manipulating data. 

Consult the cheatsheet of `list-columns` for some ideas, and this [`Jennifer Bryan presentation`](https://jennybc.github.io/purrr-tutorial/talks.html)






