---
title: "Resources"
---

Here are some resources for you to carry on after you have mastered the material on this website:

# Suggested next steps

If you would like to continue to enhance your R prowess, here is a suggested subject tree:

1. Functions - get comfortable writing your own functions for repeatable tasks.
2. Github - any coder should use some kind of source control, R's integration with Github makes it a natural fit.
3. Create a package - it doesn't have to appear in CRAN (yet!) but its a small step from sourcing a file of your custom functions to loading your own private library.
4. Shiny - being able to quickly create interactive webapps when asked for a report is always impressive 
5. Kaggle - practice your machine learning with public datasets in a friendly competative environment
6. Advanced packages - get a package on CRAN, or contribute to the development of one on Github.  Start using tests, Travis and coverage

# R and Statistics for Digital Analysts

This is a community-driven set of resources and links on R and statistics generated of, by, and for digital analysts. To submit a resource, fill out [this Google form](https://docs.google.com/a/webanalyticsdemystified.com/forms/d/e/1FAIpQLSftDmRdGT-Ad-mgaa2rEh7yt4F6hYzDGkIuIym4lsq1Lothsw/viewform). Once approved, the resource will appear here.

We will be working to improve the structure and navigation of this site. Heck...it uses Shiny, so we very well may add some interactivity, too! If you have feedback or are interested in helping out with this project, shoot a note to [Tim Wilson](mailto:tim@analyticsdemystified.com). 

```{r, echo=FALSE, message=FALSE}
library(googlesheets)
library(pander)  # To get tabs to be rendered within the code chunk

sourceBook <- gs_key("1E2H979v2L_NwrxCxI3OlfreLVTJ6a77Ed3WgH6A2YbA")

resourceList <- gs_read(sourceBook, ws = "Form responses 1")

# Pare this down to just the ones that have been approved
resourceList <- resourceList[!is.na(resourceList$`Approved?`),]

# List of resource types. The first value in each vector is the tab label. The
# Second is the value to filter on to get those resources
resourcePurposes <- data.frame(
  tabLabel = c("Tutorials","Tips & Tricks","Scripts & Packages","General Reference"),
  filterValue = c("Tutorial","Tip/Trick","Useful/Usable Script/Package","General Reference"))
  
```

```{r, echo=FALSE, results='asis'}

# The function to actually display each resource. This still needs cleanup work.
displayResource <- function(resource){
  cat(c("<p><b><a href=\"",resource$Link,"\" target='_blank'>",resource$`Resource Name/Title`,"</a></b> (",resource$`Content Type`,")<br>"),sep="")
  cat(c(resource$Description,"<br>"),sep="")
  cat(c("Content Focus: ",resource$`Focus of the Content`,"<br>"),sep="")
  if(resource$`Include Your Name as a Contributor?`=="Yes"){
    cat(c("Contributed by: ",resource$`Your Name`),sep="")
  }
  cat("</p>",sep="")
}

```

##{.tabset}

```{r, echo=FALSE, results='asis'}

for (i in 1:nrow(resourcePurposes)){
  # Add the tab and label it
  pandoc.header(as.character(resourcePurposes$tabLabel[i]),level = 3)

  # Grab the subset of resources that fit the current purpose
  resources <- resourceList[resourceList$`Resource Purpose`==as.character(resourcePurposes$filterValue[i]),]
  
  # Cycle through all of the resources and output their details
  for (j in 1:nrow(resources)){
    displayResource(resources[j,])
    }
}
```
