---
title: "Resources"
---

**Resource #1:** Join the Measure Slack team at [http://join.measure.chat](http://join.measure.chat). It's free! Be sure to join the #r-and-statistics channel!

**Resource #2:** Check out the resources below. This is a community-driven set of resources and links on R and statistics generated of, by, and for digital analysts. To submit a resource, fill out [this Google form](https://docs.google.com/a/webanalyticsdemystified.com/forms/d/e/1FAIpQLSftDmRdGT-Ad-mgaa2rEh7yt4F6hYzDGkIuIym4lsq1Lothsw/viewform). Once approved, the resource will appear here (through the magic of the `googlesheets` package and some RMarkdown!). (We are continuing to work on the structure and navigation.) 

```{r, echo=FALSE, message=FALSE}
library(googlesheets)
library(pander)  # To get tabs to be rendered within the code chunk

sourceBook <- gs_key("1E2H979v2L_NwrxCxI3OlfreLVTJ6a77Ed3WgH6A2YbA")

resourceList <- gs_read(sourceBook, ws = "Form responses 1")

# Pare this down to just the ones that have been approved
resourceList <- resourceList[!is.na(resourceList$`Approved?`),]

# List of resource types. The first value in each vector is the tab label. The
# Second is the value to filter on to get those resources
resourcePurposes <- data.frame(
  tabLabel = c("Tutorials","Tips & Tricks","Scripts & Packages","General Reference"),
  filterValue = c("Tutorial","Tip/Trick","Useful/Usable Script/Package","General Reference"))
  
```

```{r, echo=FALSE, results='asis'}

# The function to actually display each resource. This still needs cleanup work.
displayResource <- function(resource){
  cat(c("<p><b><a href=\"",resource$Link,"\" target='_blank'>",resource$`Resource Name/Title`,"</a></b> (",resource$`Content Type`,")<br>"),sep="")
  cat(c(resource$Description,"<br>"),sep="")
  cat(c("Content Focus: ",resource$`Focus of the Content`,"<br>"),sep="")
  if(resource$`Include Your Name as a Contributor?`=="Yes"){
    cat(c("Contributed by: ",resource$`Your Name`),sep="")
  }
  cat("</p>",sep="")
}

```

##{.tabset}

```{r, echo=FALSE, results='asis'}

for (i in 1:nrow(resourcePurposes)){
  # Add the tab and label it
  pandoc.header(as.character(resourcePurposes$tabLabel[i]),level = 3)

  # Grab the subset of resources that fit the current purpose
  resources <- resourceList[resourceList$`Resource Purpose`==as.character(resourcePurposes$filterValue[i]),]
  
  # Cycle through all of the resources and output their details
  for (j in 1:nrow(resources)){
    displayResource(resources[j,])
    }
}
```
