---
title: "Shiny"
output:
  html_document:
    title-prefix: "Digital Analytics R and Statistics Tutorials"
    theme: flatly
    css: styles.css
    toc: true
    toc_float: true
    includes:
      before_body: includes/analytics.html
---

```{r shiny, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
web_data <- read.csv("data/gadata_example_2.csv", stringsAsFactors = FALSE)
```

[Shiny](http://shiny.rstudio.com/) is a web application framework for R, produced by RStudio. 

A Shiny app usually has two files, `server.R` and `ui.R`.  These take care of the web server backend and the HTML frontend, respectivily. 

To run a Shiny app you need to have a Shiny server running.  RStudio comes with one pre-installed for running your apps locally, but for publishing you will need to install Shiny server or host via `shinyapps.io`.

Shiny apps use a functionality called *reactivity* that means that apps can be quick and responsive to changes to inputs - this is one of the best features of Excel, where changing one cell can have consequences throughout the Workbook.

Shiny apps can be tricky to get your head around due to its different work flow from normal R programs, its recommended to go through the [tutorials online](http://shiny.rstudio.com/tutorial/).

## Before you begin

The [Shiny website](http://shiny.rstudio.com/tutorial/) is the source of all that will be said here today, for deeper knowledge refer to that.

We will go through today some practical experience that will hopefully smooth off the learning curve, with a digital marketing focus.

## Overview

Shiny uses reactive programming.  What does that mean?

### ui.R

The `ui.R` is the equivalent of a website frontend.  It contains all the HTML, CSS and JavaScript of a normal webpage.  As such its usually easier to get to grips with.

All the most HTML tags are available using `shiny::tags()`, with shortcuts to the most common e.g. `div()`

The most minimal UI you can make looks like this:

```r
## in ui.R
shinyUI(
  fluidPage(
    ## output functions
)
```

The functions `shinyUI` and `fluidPage` create the minimal HTML necessary for a Shiny app:

```{r}
library(shiny)
shinyUI(fluidPage())
```

You will need to deal with *a lot* of brackets.  Be stringent in knowing which closing brack refers to which function, as it can be very hard to debug.  Good code style indentation helps:

```r
library(shiny)
shinyUI(
  fluidPage(
    
  ) ## end fluidPage
) ## end shinyUI
```

Building a UI is then a matter of building up the structure in lots of nested UI functions. 

#### Interaction with outputs 

Most UI elements that will work with the `server.R` are called `blahOutput()`.  

Behind the scenes this sets up a div tag with the ID of the element that will be changed dynamically via `server.R`:

```r
## in ui.R
plotOutput("example")
## <div id="example" class="shiny-plot-output" style="width: 100% ; height: 400px"></div>
```

`server.R` interacts with this div ID via a named lists called `output` - *it outputs changes to this ID*.  If the ID is not there, you will see nothing. 

The ID must match the list name, and the type must be the same via the `renderBlah()` server functions e.g. `plotOutput` in `ui.R`, `renderPlot` in `server.R`

```r
## in server.R
output$example <- renderPlot({
  ## a function that creates a plot
  plot(mtcars)
})
```

The above uses `output$example` - this will output to `plotOutput("example")`

#### Working with inputs

Likewise, all the input IDs are available to `server.R` via a named list called `input`.

The input elements in `ui.R` are all called something like `blahInput` e.g. `sliderInput()`

Again, you specify an ID for that input, but also criteria on what a user can input:

```r
library(shiny)

## create a slide input
sliderInput("input1", label = "A slider", min = 1, max = 10, value = 2)

# <div class="form-group shiny-input-container">
#  <label class="control-label" for="input1">A slider</label>
#  <input class="js-range-slider" id="input1" data-min="1" data-max="10" data-from="2" data-step="1" 
# data-grid="true" data-grid-num="9" data-grid-snap="false" data-prettify-separator="," data-keyboard="true" 
# data-keyboard-step="11.1111111111111" data-drag-interval="true" data-data-type="number"/>
# </div>
```

The output is more complicated than the `blahOutput` functions, but again it is only HTML. 

To refer to the input elements, `server.R` then refers to a named list called `input`:

```r
input$input1
```

Inputs effect outputs, and hence we get interactivity.

A very basic minimal example below:

```r
## ui.R
library(shiny)

shinyUI(
  fluidPage(
    textInput("text_input", label = "input text"),
    textOutput("text_output")
  )
)
```

```r
# server.R
library(shiny)

function(input, output) ({
   
  output$text_output <- renderText({
    input$text_input
  })
  
})

```

All that said, its best to start with `ui.R` first to design your UI, then work on the `server.R` to output what you need. 

### server.R

`server.R` contains the code that will change certain elements defined in the `ui.R`.

`server.R` is in fact a constantly looping function.  As a user interacts with the `ui.R` it sends messages back to the `server.R` which then reacts (see, `reactivity`), outputs data to change and sends this back to the `ui.R` that renders it on the correct ID. 

The looping function is around the objects we spoke about earlier - `input` and `output`.  Optionally, you can also include an object called `session` which contains information about the general user session (such as what URL is being shown), which starts to explain how a minimal `server.R` file looks:

```r
## in server.R
library(shiny)

function(input, output){

  output$plot1 <- renderPlot({
  
    ## get the data from csv
    web_data <- read.csv("gadata_example_2.csv", stringsAsFactors = FALSE)
    plot(web_data$date, web_data[[input$value1]])
  
  })

}
```

## Exercise

1. What will this Shiny app render?  Assume input$value1 is the name of a metric column of `web_data`

2. Make a Shiny app in a new RStudio project. (Click `New > Shiny > Multiple File` in RStudio interface)  construct the ui.R that will alter the `server.R` code above.  You will need to use the functions `selectInput()` to choose the data column you want to plot and `plotOutput()` to output the result.

3. What steps should we take to use highcharter instead of base plot? 

Use these functions - `renderHighchart`, `highchartOutput` and to make the plot with the below:

```r
library(highcharter)
hchart(web_data, "line", hcaes(x = date, y = sessions))
```
 
## Reactive objects

Because its a looping function, standard R rules about scoping don't apply.  You need to deal with functions like `reactive()`, `observe()`,`isolate()`, `eventReactive()` and `observeEvent()` which control behaviour in this looping function. 

Key to this are a new class of object called `reactive` objects. These are objects that will change according to user input, and as such can't be treated as normal R variables.

Reactive object can react to another reactive object, which sets up a chain of dependencies.  A change in a dependency will effect all other reactive objects.  A key factor in making efficient Shiny apps is to know when to trigger a change and when not to.

### Example

The below builds on the app below, but we now also want to display the maximum value.  As we will now have two UI elements we will have two server.R calls that rely on the same data.  

Instead of importing the data in twice (it could be a API call) we will make a reactive object, and then when that object is ready (e.g. not `NULL`) pass it to the functions.  

The `req()`function serves as a roadblock to stop the plot and text firing before the data is ready (important for API calls)

We use `highcharter` as the plots are cool.

```r
## in server.R
library(shiny)
library(highcharts)

function(input, output){
  
  ## our reactive object for data
  the_data <- reactive({
    
    ## get the data from csv
    the_data <- read.csv("../data/gadata_example_2.csv", stringsAsFactors = FALSE)
    
    ## return the data with the date column for the plot and metric column selected
    the_data[ , c("date", input$metric)]
    
  })
  
  output$plot1 <- renderHighchart({
    
    ## this is a roadblock so this function won't fire until the_data is non-NULL
    req(the_data())
    
    ## do this as it make debugging a lot easier
    plot_data <- the_data()
    
    ## creates a line chart using highcharts
    hchart(plot_data, "line" , hcaes(x = date)) %>% 
      hc_add_series(plot_data[[input$metric]])
    
  })
  
  output$max_text <- renderText({
    
    ## this is a roadblock so this function won't fire until the_data is non-NULL
    req(the_data())
    req(input$metric)
    
    ## do this as it make debugging a lot easier
    max_data <- the_data()
    
    paste("The max is: ", max(max_data))
    
  })
  
}
```

The `req(input$value1)` is good practice to have as the `input list` are not necessarily available to the function on start up - this guards against ugly red messages briefly when the app starts up.

With a ui.R adding the max text:

```r
## in ui.R

library(shiny)
library(highcharter)

shinyUI(
  fluidPage(
    selectInput("metric", label = "select column", choices = c("sessions",
                                                               "pageviews",
                                                               "entrances",
                                                               "bounces")),
    textOutput("max_text"),
    highchartOutput("plot1")
  )
)
```

### Exercise

1. Add a `selectInput` that also lets you choose the `deviceCategory`

> HINT: you may have a line similar to `plot_data <- plot_data[plot_data$device == input$device, c("date", input$metric)]` to segment your data (or maybe not....check your data and put a breakpoint in the correct place)

Use `dplyr` to massage the data if necessary

2. Extra marks - add another `selectInput` with `channelGrouping` - this may be useful:

```r
c("(Other)","Direct","Display","Email","Organic Search","Paid Search","Referral","Social","Video")
```


## Shiny Modules

Its possible to package up common UI elements into a [Shiny module](https://shiny.rstudio.com/articles/modules.html).  This can let you repeat templated code. 

An example comes with `googleAnalyticsR` to let you create a login for users.  The below is from the documentation.

```r
## in server.R
library(googleAuthR)
library(googleAnalyticsR)
library(shiny)
library(highcharter)

function(input, output, session){
  
  ## Get auth code from return URL
  token <- callModule(googleAuth, "login")
  
  ga_accounts <- reactive({
    req(token())
    
    with_shiny(ga_account_list, shiny_access_token = token())
    
  })
  
  ## get the ViewId from what the users select in the dropdown
  selected_id <- callModule(authDropdown, "auth_menu", ga.table = ga_accounts)
  
  gadata <- reactive({
    
    req(selected_id())
    gaid <- selected_id()
    
    ## with shiny needed to read the correct token
    with_shiny(google_analytics,
               id = gaid,
               start=input$date_select[1], end=input$date_select[2], 
               metrics = c("sessions"), 
               dimensions = c("date"),
               shiny_access_token = token())
  })
  
  output$something <- renderHighchart({
    
    ## only trigger once authenticated
    req(gadata())
    
    gadata <- gadata()
    
    ## creates a line chart using highcharts
    hchart(gadata, "line" , hcaes(x = date, y = sessions))
    
  })
  
}

```

The modules are used below to render the dropdowns and login buttons:

```r
## ui.R
library(googleAuthR)
library(googleAnalyticsR)
library(shiny)
library(highcharter)

shinyUI(
  fluidPage(
    googleAuthUI("login"),
    authDropdownUI("auth_menu"),
    dateRangeInput("date_select", "Select Date", start = Sys.Date() - 30),
    highchartOutput("something")
    
  ))
```




